Building an SSI Cluter with GNU/Hurd

### 2019-05-31 libpager/paper.tex
finishing up "Building an SSI cluster on Hurd"

Hurd's \netmsg server and the \libpager library have been developed
to the point where Hurd could be usable as a cluster operating system,
though the current Debian distribution doesn't use this functionality.
At the time of this writing (2019), the major barriers to building a
Hurd cluster are the 32-bit address space and the lack of SMP support,
enhancements that are beyond the scope of the work described in this
paper.

@@

\begin{figure}
\begin{center}
\begin{tikzpicture}[node distance = .5cm]

@@

This is basically the only way in which a process can obtain Mach
ports since, much like POSIX file descriptors, port numbers are local
to a process and interpreted by the kernel, so there is no way that a
process can fabricate another process's port number, just as a POSIX
process can not fabricate another process's file descriptor.  A
process is provided a few standard ports when it is initialized, one
of which being a port with send rights that represents the root of the
file system.  By sending filename lookup requests to this port, the
process can traverse the entire file system tree, each node of which
is represented by a different port.  The hurd file system is thus a
hierarchy of ports, most of which recognize a standard set of RPCs
performing file operations.  The primary way for a program to
advertise its services in Hurd is for it to listen for messages on a
port that it attachs to a name in the file system.

@@

For example, consider how Hurd handles an {\tt mmap()} call.  A
program uses directory lookup calls to obtain a file port, which is a
send right to a port managed by a file server, and corresponds to a
standard POSIX file descriptor.  It then uses the {\tt io_map} RPC to
obtain another send right, also to a port managed by the file server,
which implements the Mach memory management protocol.  It passes that
send right to the kernel, using a {\tt vm_map} RPC, indicating that it
wishes to map that memory object (i.e, the file), into its memory
space.  The kernel, after a \moinit / \moready handshake with the file
server, creates a new memory region in the program's address space,
initially with no physical memory assigned to it, and returns that
address to the program.

As soon as the program attempts to access this region of memory in any
way, a page fault is triggered, which causes the kernel to send an
\morequest message to the file server.  The file server responds by
reading the required page from disk (if necessary) and passing it to
the kernel by attaching it to a \mosupply message.  The kernel, now
managing this region of memory, maps it to the correct page in the
previously assigned region of memory, and allows the program to
continue execution.

@@

\begin{center}
\begin{tikzpicture}

@@

\draw [color=red, <-, thick] (2 em,1) -- node [midway, above] {io\_map} (13 em,1);
\draw [color=red, ->, thick, dotted] (2 em,.9) -- node [midway, below] {\tiny port 153} (13 em,.9);

\draw [color=red, ->, thick] (15 em,-3 em) -- node [midway, left, align=center] {vm\_map\\{\tiny port 153}} (15 em,- 7 em);
\draw [color=red, <-, thick, dotted] (15.2 em,-3 em) -- node [midway, right] {\tiny addr} (15.2 em,- 7 em);

\draw [color=red, <-, thick] (-2 em,-3 em) -- node [midway, left] {m\_o\_init} (-2 em,- 7 em);
\draw [color=red, ->, thick] (-1.5 em,-3 em) -- node [midway, right] {m\_o\_ready} (-1.5 em,- 7 em);

%\draw [color=red, <-, thick] (0.5 em,-3 em) -- node [midway, left] {m\_o\_data\_request} (0.5 em,- 8 em);
%  \draw [color=red, ->, thick] (1 em,-3 em) -- node [midway, right] {m\_o\_data\_supply} (1 em,- 8 em);

\end{tikzpicture}
\end{center}

@@

Several \libpager implementation choices exist, and perhaps should be configuration options.

@@

\begin{verbatim}
Global structure {
   /* NEXTERROR list
    * A linked list of outstanding (error code, page, client) tuples.
    * They'll be transmitted to the client that requested them.
    */

   /* WRITEWAIT list
    * A linked list of pages waiting to be written to backing store
    */

    std::map<std::pair<vm_offset_t, vm_size_t>, outstanding_lock> outstanding_locks;

    std::set<outstanding_change_request> outstanding_change_requests;
}
\end{verbatim}

@@

Another possible bug scenario (``bug3''): a single client holds two
pages with write access.  It returns one (perhaps because \pagersync
was called) which triggers a call to \pagerwritepage.  While
\pagerwritepage is running, both pages are returned in a single
\moreturn message.  Since the first page is still writing, we
sleep to wait for it to finish.  Then the second page is requested
again.  We {\it should} serve the copy that's queued to write, but if
we're not careful we'll call \pagerreadpage on the second page.

>>>>

Hurd's \netmsg server and the \libpager library have been developed
to the point where Hurd could be usable as a cluster operating system,
though the current Debian distribution doesn't use this functionality.
At the time of this writing (2019), the major barriers to building a
Hurd cluster are the 32-bit address space and the lack of SMP support.

@@

\begin{figure}[ht]
\begin{center}
\begin{tikzpicture}[node distance = .5cm]

@@

This is basically the only way in which a process can obtain Mach
ports since, much like POSIX file descriptors, port numbers are local
to a process and interpreted by the kernel, so there is no way that a
process can fabricate another process's port number, just as a POSIX
process can not fabricate another process's file descriptor.  A
process is provided a few standard ports when it is initialized, one
of which being a port with send rights that represents the root of the
file system.  By sending filename lookup requests to this port, the
process can traverse the entire file system tree, each node of which
is represented by a different port.  The hurd file system is thus a
hierarchy of ports, most of which recognize a standard set of RPCs
performing file operations.  The primary way for a program to
advertise its services in Hurd is for it to listen for messages on a
port that it attaches to a name in the file system.

@@

For example, consider how Hurd handles an {\tt mmap()} call (Figure
\ref{Initial mapping of a Mach memory object}).  A program uses
directory lookup calls to obtain a file port, which is a send right to
a port managed by a file server, and corresponds to a standard POSIX
file descriptor.  It then uses the {\tt io_map} RPC to obtain another
send right, also to a port managed by the file server, which
implements the Mach memory management protocol.  It passes that send
right to the kernel, using a {\tt vm_map} RPC, indicating that it
wishes to map that memory object (i.e, the file), into its memory
space.  The kernel, after a \moinit / \moready handshake with the file
server, creates a new memory region in the program's address space,
initially with no physical memory assigned to it, and returns that
address to the program.

As soon as the program attempts to access this region of memory in any
way (Figure \ref{Initial access of a Mach memory object}), a page
fault is triggered, which causes the kernel to send an \morequest
message to the file server.  The file server responds by reading the
required page from disk (if necessary) and passing it to the kernel by
attaching it to a \mosupply message.  The kernel, now managing this
region of memory, maps it to the correct page in the previously
assigned region of memory, and allows the program to continue
execution.

@@

\tikzstyle{message} = []

\begin{figure}
\begin{center}
\begin{tikzpicture}

@@

\draw [message, <-, thick] (2 em,1) -- node [midway, above] {io\_map} (13 em,1);
\draw [message, ->, thick] (2 em,.9) -- node [midway, below] {\tiny port 153} (13 em,.9);

\draw [message, ->, thick] (15 em,-3 em) -- (15 em,- 7 em);
\draw [message, <-, thick] (15.2 em,-3 em) -- node [midway, right] {\tiny addr} (15.2 em,- 6 em);

\node [message, left, align=center] at (15 em, - 4 em) {vm\_map\\{\tiny port 153}};

\draw [message, <-, thick] (-2 em,-3 em) -- (-2 em,- 7 em);
\draw [message, ->, thick] (-1.5 em,-3 em) -- (-1.5 em,- 7 em);

\node [message, left] at (-2 em, -5.6 em) {m\_o\_init};
\node [message, right] at (-1.5 em, -5.6 em) {m_o_ready};

\node (port) at (-2 em, -1.5 em) [minimum size=1.5cm, draw, thick] {Port 153};

%\draw [message, <-, thick] (0.5 em,-3 em) -- node [midway, left] {m\_o\_data\_request} (0.5 em,- 8 em);
%  \draw [message, ->, thick] (1 em,-3 em) -- node [midway, right] {m\_o\_data\_supply} (1 em,- 8 em);

\end{tikzpicture}
\end{center}
\caption{Initial mapping of a Mach memory object}
\label{Initial mapping of a Mach memory object}
\end{figure}

\begin{figure}
\begin{center}
\begin{tikzpicture}

\draw [message, <-, thick] (-2 em,-3 em) -- (-2 em,- 7 em);
\draw [message, ->, thick] (-1.5 em,-3 em) -- (-1.5 em,- 7 em);

\node [message, left] at (-2 em, -5.6 em) {m\_o\_data_request};
\node [message, right] at (-1.5 em, -5.6 em) {m_o_data_supply};

%%\draw [message, ->, thick, dotted] (memory2.south) -- node [midway, right] {fault} (memory2.south |- kernel);

\node (port) at (-2 em, -1.5 em) [minimum size=1.5cm, draw, thick] {Port 153};

\node (instr) [right, align=center, font=\tiny] at (Program -| Program.west) {INSTR\\ READ\\ INSTR};
\node [draw] (location) at (Program) {};
\draw [->] (instr.east) -- (location.west);

\draw [color=red, ->, thick, dotted] (location.south) -- node [midway, right] {fault} (location.south |- kernel);

%\draw [message, <-, thick] (0.5 em,-3 em) -- node [midway, left] {m\_o\_data\_request} (0.5 em,- 8 em);
%  \draw [message, ->, thick] (1 em,-3 em) -- node [midway, right] {m\_o\_data\_supply} (1 em,- 8 em);

\end{tikzpicture}
\end{center}
\caption{Initial access of a Mach memory object}
\label{Initial access of a Mach memory object}
\end{figure}

@@

Several \libpager implementation choices exist, and perhaps should be configuration options,
though they are currently hard-wired into the code.

@@

The pagemap is organized as an array with one entry for each virtual
memory page in the memory object.  In addition to the pagemap, there
are several additional data structures maintained globally (i.e, for
the entire memory object):

\begin{itemize}[noitemsep]
\item The NEXTERROR list, a linked list of outstanding (error code, page, client) tuples
\item A list of pages waiting to be written to backing store
\item A list of outstanding lock requests
\item A list of outstanding change requests
\end{itemize}

@@

Another possible bug scenario runs like this: a single client holds two
pages with write access.  It returns one (perhaps because \pagersync
was called) which triggers a call to \pagerwritepage.  While
\pagerwritepage is running, both pages are returned in a single
\moreturn message.  Since the first page is still writing, we
sleep to wait for it to finish.  Then the second page is requested
again.  We {\it should} serve the copy that's queued to write, but if
we're not careful we'll call \pagerreadpage on the second page.
